02_models_1

### 2025-04-07 02_models_1.Rmd
added creating folders, runs models with all species and only with temp

---
title: "Models, derivatives, response diversity"
author: "Sophie Moreau"
date: "2024-07-05"
output: html_document
editor_options: 
  chunk_output_type: console
---
#General Additive Models
Running the predictions for all GAMs with the smaller data frames to reduce running 
time.

@@

###Predictions of GAMs

```{r}
# read in the dataframes
df_1 <- readRDS("data_frame_models/df_binomial_gam")
df_2 <- readRDS("data_frame_models/df_abundance_gam")
df_3 <- readRDS("data_frame_models/df_binomial_re")
df_4 <- readRDS("data_frame_models/df_abundance_re")

# we have three different functions. One predictions() does not simply any models
# predictions_temp() has multiple models that are simplified (see supplementary 
# figures for the disucssion), predictions_depth_temp() models all species with 
# abundance > 20 with both coviarates temperature and depth.

# For downstream analysis for thermal response diversity models from predictions_temp()
# are used.

# temperature models with simplified models 
# predictions take VERY LONG! and two folders are needed to store the 
# DHARMa and prediction output inside: 
# total_models/gam_check
# needs a folder called total_models/predictions/temp

predictions_temp(df_1)
predictions_temp(df_2)
predictions_temp(df_3)
predictions_temp(df_4)

simplified_predictions_all <- list.files(path = "total_models/predictions/temp", pattern = ".rds", full.names = TRUE) |>
  map_dfr(readRDS)

# save total predictions as RDS
saveRDS(simplified_predictions_all, "total_models/df_pred_all.rds")

# you can directly use the dataframe df_pred_all for downstream analysis

```
###Simplification and temp depth models
```{r}
# both other functions:

# you can also directly use the dataframe df_pred_no_simplification for decisions about
# which models should be simplified 
# and
# depth_temp_pred_all for the sensitivity testing of the temp + depth models

# testing for which models need to be simplified
# needs a folder: total_models/predictions/temp_test

# all temperature models, no simplificaiton
# predictions(df_1)
# predictions(df_2)
# predictions(df_3)
# predictions(df_4)
# 
# no_exclusion_predictions_all <- list.files(path = "total_models/predictions/temp_test", pattern = ".rds", full.names = TRUE) |>
#   map_dfr(readRDS)
# 
# # save total predictions as RDS
# saveRDS(no_exclusion_predictions_all, "total_models/df_pred_no_simplification.rds")

# testing if modelling temp and depth improves the models
# needs a folder: total_models/predictions/depth_temp

# sensitivity
# predictions_depth_temp(df_1)
# predictions_depth_temp(df_2)
# predictions_depth_temp(df_3)
# predictions_depth_temp(df_4)
# 
# # issue with cottus sp po

# # statistics of the gamms with dept and temp
# total_models <- bind_rows(df_1, df_2, df_3, df_4)
# 
# stats_gamms(total_models)
# 
# #  compile data frame with all predictions from the saved species' model predictions
# 
# depth_temp_predictions_all <- list.files(path = "total_models/predictions/depth_temp", pattern = ".rds", full.names = TRUE) |>
#   map_dfr(readRDS)
# 
# # save total predictions as RDS
# saveRDS(depth_temp_predictions_all, "total_models/depth_temp_pred_all.rds")

```

####Fig. 2b
New figure 2b with categories across all lakes

```{r, fig.show='hide'}

# Load total predictions
model_predictions <- readRDS("total_models/df_pred_all.rds")

# check the species list
model_predictions$species <- as.factor(model_predictions$species)
levels(model_predictions$species)

# read in the category table and rename categories
species_across_lakes <-  read_xlsx("species_category_across_lakes.xlsx") |> 
  select(-notes) |> 
    mutate(category = ifelse(category == "native", "non_endemic_native", category)) |>
  mutate(category2 = factor("all"))

species_across_lakes$category <- as.factor(species_across_lakes$category)
levels(species_across_lakes$category)

#merge predictions with categories
predictions_categories <- merge(model_predictions, species_across_lakes)

# change category names for facet_wrap
category_names <- c(
  `endemic` = "endemic",
  `non_endemic_native` = "non-endemic native",
  `non_native` = "non-native"
)

# colors for the plot
mycolors1 <-  c("endemic"= "#63D8F2", "non_endemic_native"="#F26379", "non_native" = "#5A64D8")

# Rescale the predictions to 0-1 
predictions_categories_rescaled <- predictions_categories |>
  group_by(species) |>
  mutate(fit_rescaled = (fit - min(fit)) / (max(fit) - min(fit)))

plot_2b <- predictions_categories_rescaled |>
  ggplot(aes(temp, fit_rescaled, group = species, color = category)) +
   geom_line(linewidth = 0.5) +
  theme_bw(base_size = 16) +
  ylab("Abundance") +
  xlab("Temperature") +
  scale_color_manual(values = mycolors1, guide = NULL) +
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(), 
        legend.position = "bottom",
        legend.title = element_blank())

plot_2b_final <- plot_2b + facet_grid(~category, labeller = as_labeller(category_names))

plot_2b_final

# save plot as TIFF
# tiff(paste("total_models/plots/plot_2b.tiff", sep = ""), units="in", width=10, height=4, res=300)
# 
# plot(plot_2b_final)
# 
# dev.off()

```

####Fig. Supplementary predictions plot
```{r}
# also using predictions_categories from chunk above
# trying to group by category and color

str(categories_supp)

categories_supp <- predictions_categories |>
  arrange(category) |> 
  mutate(species = fct_reorder(species, as.numeric(category))) |> 
  ggplot(aes(temp, fit)) +
  geom_ribbon(aes(ymin = (fit - se.fit), ymax = (fit + se.fit), fill = factor(category)), alpha = 0.3) +
  geom_line(aes(color = factor(category))) +
  theme_bw() +
  facet_wrap(~species, scale = "free_y", labeller = labeller(species = label_wrap_gen(width = 10))) +
  theme(strip.background = element_rect(fill = "lightgrey"), strip.text = element_text(size = 6)) +
  scale_color_manual(values = mycolors1, aesthetics = c("color", "fill")) +
  xlab("Temperature") +
  ylab("Abundance")

categories_supp

# tiff(paste("total_models/plots/supplementary_predictions.tiff", sep = ""), units="in", width = 12, height=8, res=300)
# plot(categories_supp)
# 
# # Closing the graphical device
# dev.off()

```
##Sensitivity testing
###NEW Fig. S1
Testing if depth-driven temperature models can adequately fit abundance data compared to a temperature and depth model.

```{r}
temp_only <- readRDS("total_models/df_pred_all.rds") |> 
  mutate(model = "temp")

temp_depth <- readRDS("total_models/depth_temp_pred_all.rds") |> 
  mutate(model = "temp_depth")

sensitivity_predictions <- bind_rows(temp_only, temp_depth)

supplementary_sensitivity <- sensitivity_predictions |>
    ggplot(aes(temp, fit, color = model, fill = model)) +
    geom_line() +
    # geom_ribbon(aes(ymin = (fit_rescaled - se.fit), ymax = (fit_rescaled + se.fit)), color = NA, alpha = 0.1) +
    theme_bw() +
    # labs(title = paste("Sensitivity Test", i)) +
    scale_color_manual(values = c("#D73027", "#08306B"), aesthetics = c("color", "fill")) +
     facet_wrap(~species, scale = "free", labeller = labeller(species = label_wrap_gen(width = 10))) +
    theme(strip.background = element_rect(fill = "lightgrey"), strip.text = element_text(size = 6)) +
    ylab("Abundance") +
    xlab("Temperature") +
   theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(), 
        legend.position = "right",
        legend.title = element_blank())

# # save the plot as TIFF
# tiff(paste("total_models/plots/supplementary_sensitivity_predictions.tiff", sep = ""), units="in", width = 12, height=8, res=300)
# plot(supplementary_sensitivity)
# 
# # Closing the graphical device
# dev.off()

```
####AIC comparison 
```{r}
# to be added
```

#Derivatives calculation

##Data preparation
We calculated derivatives for each species per lake.

```{r}

# Derivatives() will ignore any random effect smooths it encounters in object.

# Table of pecies that were recorded with electro, fishbase or eawag, following Table 16 in Projet Lac Synthesis Report. 
# Plus the minimum and maximum temp in each lake -> needed for the new data for 
# calculating derivatives

species_lake <- read_xlsx("species_occurrences_lakes.xlsx") 
str(species_lake)

species_lake$fLake <- as.factor(species_lake$Lake)
species_lake$fProtocol <- as.factor(species_lake$Protocol)
species_lake$Species <- as.factor(species_lake$Species)

#Loading all four dfs to calculate derivatives

df_binomial_gam <- readRDS("data_frame_models/df_binomial_gam")
df_binomial_gam$fLake <- as.factor(df_binomial_gam$Lake)
df_binomial_gam$fProtocol <- as.factor(df_binomial_gam$Protocol)

df_abundance_gam <- readRDS("data_frame_models/df_abundance_gam")
df_abundance_gam$fProtocol <- as.factor(df_abundance_gam$Protocol)
df_abundance_gam$fLake <- as.factor(df_abundance_gam$Lake)

df_binomial_re <- readRDS("data_frame_models/df_binomial_re")
df_binomial_re$fLake <- as.factor(df_binomial_re$Lake)
df_binomial_re$fProtocol <- as.factor(df_binomial_re$Protocol)

df_abundance_re <- readRDS("data_frame_models/df_abundance_re")
df_abundance_re$fLake <- as.factor(df_abundance_re$Lake)
df_abundance_re$fProtocol <- as.factor(df_abundance_re$Protocol)

```

##Derivatives
```{r}

# Now four different loops are run to calculate derivatives of each species 
# in each lake

# Model 1: Binomial data without random intercepts

species_list <- df_binomial_gam |>
  distinct(Species) |>
  pull(Species)

species_list <- sort(species_list)

derivatives <- list()
gam_output <- list()
model_prediction <- list()

# loop through the list of species
for (i in species_list) {
  data <- df_binomial_gam |>
    filter(Species == i)

  lake_data <- species_lake |>
    filter(Species == i)

  gam_output[[i]] <- gam(data = data, Abundance ~ s(mean_last_7days, k = 3) + s(fProtocol, bs = 're'), family = binomial)

  lake_list <- distinct(lake_data, Lake) |>
    pull()

  # loop through all lakes where the species is present, lake list is based on the information from
  # Projet Lac
  for (j in lake_list){

    data_lake <- species_lake |>
      filter(Species == i) |>
      filter(Lake == j)

    # using minimum and maximum temp in the lake for derivatives
     newdata <- tibble(mean_last_7days = seq(from = min(data_lake$temp, na.rm = TRUE),
                                             to = max(data_lake$temp, na.rm = TRUE), length = 200),
                       fProtocol = factor("VERT"))

     # calculate derivatives
     derivatives <- derivatives(gam_output[[i]], data = newdata) |>
       mutate(fLake = factor(j)) |>
       mutate(species = factor(i)) |>
       rename(temp = mean_last_7days)

     saveRDS(derivatives, paste0("total_models/derivatives/derivatives_", i, "_",  j, ".rds"))
  }
}

# Model 2.1: Abundance data and no random intercept for lake -> but binomial
# Coregonus_sp_benthic_profundal did not run with ZIP and was thus
# run with binomial.

species_list <- df_abundance_gam |> 
  # binomial one
  filter(Species == "Coregonus_sp_benthic_profundal") |>
  distinct(Species) |> 
  pull(Species)

species_list <- sort(species_list)

derivatives <- list()
gam_output <- list()
model_prediction <- list()

for (i in species_list) {

  data <- df_abundance_gam |> 
    filter(Species == i)

  # binomial family instead of ZIP
  gam_output[[i]] <- gam(data = data, Presence ~ s(mean_last_7days, k = 3) +
                           s(fProtocol, bs = 're'), family = binomial())

  lake_data <- species_lake |>
    filter(Species == i)

  lake_list <- distinct(lake_data, Lake) |>
    pull()

  for (j in lake_list){

    data_lake <- species_lake |>
      filter(Species == i) |>
      filter(Lake == j)

    newdata <- tibble(mean_last_7days = seq(
      from = min(data_lake$temp, na.rm = TRUE),
      to = max(data_lake$temp, na.rm = TRUE), length = 200),
      fProtocol = factor("VERT"))

    derivatives <- derivatives(gam_output[[i]], data = newdata) |>
      mutate(fLake = factor(j)) |>
      mutate(species = factor(i)) |> 
      rename(temp = mean_last_7days)

    saveRDS(derivatives, paste0("total_models/derivatives/derivatives_", i, "_",  j, ".rds"))
  }
}

# Model 2.2: Abundance data and no random intercept for lake -> ZIP
# all other species in this group run with ZIP

species_list <- df_abundance_gam |>
  filter(!Species == "Coregonus_sp_benthic_profundal") |>
  distinct(Species) |> 
  pull(Species)

species_list <- sort(species_list)

derivatives <- list()
gam_output <- list()
model_prediction <- list()

for (i in species_list) {

  data <- df_abundance_gam |> 
    filter(Species == i)

  gam_output[[i]] <- gam(data = data, Abundance ~ s(mean_last_7days, k = 3) + s(fProtocol, bs = 're'),
                         family = ziP())

  lake_data <- species_lake |>
    filter(Species == i)

  lake_list <- distinct(lake_data, Lake) |>
    pull()

  for (j in lake_list){

    data_lake <- species_lake |>
      filter(Species == i) |>
      filter(Lake == j)

    newdata <- tibble(mean_last_7days = seq(
      from = min(data_lake$temp, na.rm = TRUE),
      to = max(data_lake$temp, na.rm = TRUE), length = 200),
      fProtocol = factor("VERT"))

    derivatives <- derivatives(gam_output[[i]], data = newdata) |>
      mutate(fLake = factor(j)) |>
      mutate(species = factor(i)) |>
      rename(temp = mean_last_7days)

    saveRDS(derivatives, paste0("total_models/derivatives/derivatives_", i, "_",  j, ".rds"))
  }
}

# Model 3: Binomial data and random intercept for lake

species_list <- df_binomial_re |> 
  distinct(Species) |> 
  pull(Species)

species_list <- sort(species_list)

derivatives <- list()
gam_output <- list()
model_prediction <- list()

for (i in species_list) {

  data <- df_binomial_re |> 
    filter(Species == i)

  gam_output[[i]] <- gam(data = data, Abundance ~ s(mean_last_7days, k = 3) + s(fLake, bs = "re")
                         + s(fProtocol, bs = 're'), family = binomial)

  lake_data <- species_lake |>
    filter(Species == i)

  lake_list <- distinct(lake_data, Lake) |>
    pull()

  unique_lakes <- distinct(data, Lake) |> 
    pull()

  for (j in lake_list){

    data_lake <- species_lake |>
      filter(Species == i) |>
      filter(Lake == j)

    # take all lakes where the species is present and take random one for 
    # the data included in derivatives()
    random_lake <- sample(unique_lakes, 1)

    newdata <- tibble(mean_last_7days = seq(
      from = min(data_lake$temp, na.rm = TRUE),
      to = max(data_lake$temp, na.rm = TRUE), length = 200),
      fProtocol = factor("VERT"), fLake = factor(random_lake))

    derivatives <- derivatives(gam_output[[i]], data = newdata) |>
      mutate(fLake = factor(j)) |>
      mutate(species = factor(i)) |>
      rename(temp = mean_last_7days)

    saveRDS(derivatives, paste0("total_models/derivatives/derivatives_", i, "_",  j, ".rds"))

  }
}

# Model 4.1: Abundance data and random intercept for lake, simplified models
# multiple species in this group were simplified and run with binomial

species_list <- df_abundance_re |>
  # simplified models
  filter(Species %in% c("Alburnus_arborella", "Barbatula_sp_Lineage_I",
                        "Cyprinus_carpio","Phoxinus_csikii", "Salmo_trutta")) |>
  distinct(Species) |>
  pull(Species)

species_list <- sort(species_list)

derivatives <- list()
gam_output <- list()
model_prediction <- list()

for (i in species_list) {

  data <- df_abundance_re |>
    filter(Species == i)

  gam_output[[i]] <- gam(data = data, Presence ~ s(mean_last_7days, k = 3) + s(fLake, bs = 're')
                         +  s(fProtocol, bs = 're'), family = binomial)

  lake_data <- species_lake |>
    filter(Species == i)

  lake_list <- distinct(lake_data, Lake) |>
    pull()

  unique_lakes <- distinct(data, Lake) |>
    pull()

  for (j in lake_list){

    data_lake <- species_lake |>
      filter(Species == i) |>
      filter(Lake == j)

    random_lake <- sample(unique_lakes, 1)

    newdata <- tibble(mean_last_7days = seq(
      from = min(data_lake$temp, na.rm = TRUE),
      to = max(data_lake$temp, na.rm = TRUE), length = 200),
      fLake = factor(random_lake), fProtocol = factor("VERT"))

    derivatives <- derivatives(gam_output[[i]], data = newdata) |>
      mutate(fLake = factor(j)) |>
      mutate(species = factor(i)) |>
      rename(temp = mean_last_7days)
    saveRDS(derivatives, paste0("total_models/derivatives/derivatives_", i, "_",  j, ".rds"))

  }
}

# Model 4.1: Abundance data and random intercept for lake, complex models

species_list <- df_abundance_re |> 
  # exclusion of simplified models
  filter(!Species %in% c("Alburnus_arborella", "Barbatula_sp_Lineage_I",
                        "Cyprinus_carpio", "Phoxinus_csikii", "Salmo_trutta")) |>
  distinct(Species) |> 
  pull(Species)

species_list <- sort(species_list)

derivatives <- list()
gam_output <- list()
model_prediction <- list()

for (i in species_list) {

  data <- df_abundance_re |> 
    filter(Species == i)

  gam_output[[i]] <- gam(data = data, Abundance ~ s(mean_last_7days, k = 3) + s(fLake, bs = 're')
                         +  s(fProtocol, bs = 're'), family = ziP())

  lake_data <- species_lake |>
    filter(Species == i)

  lake_list <- distinct(lake_data, Lake) |>
    pull()

  unique_lakes <- distinct(data, Lake) |> 
    pull()

  for (j in lake_list){

    data_lake <- species_lake |>
      filter(Species == i) |>
      filter(Lake == j)

    random_lake <- sample(unique_lakes, 1)

    newdata <- tibble(mean_last_7days = seq(
      from = min(data_lake$temp, na.rm = TRUE),
      to = max(data_lake$temp, na.rm = TRUE), length = 200),
      fLake = factor(random_lake), fProtocol = factor("VERT"))

    derivatives <- derivatives(gam_output[[i]], data = newdata) |>
      mutate(fLake = factor(j)) |>
      mutate(species = factor(i)) |>
      rename(temp = mean_last_7days)

    saveRDS(derivatives, paste0("total_models/derivatives/derivatives_", i, "_",  j, ".rds"))

  }
}

# save all the derivatives in one file
# 
df_deriv <- list.files(path = "total_models/derivatives", pattern = ".rds", full.names = TRUE) |>
  map_dfr(readRDS)

saveRDS(df_deriv, "total_models/df_deriv_all.rds")

```

#Testing derivatives
##Simplification of models
```{r}
# to get the derivatives before simplifying any model: 
# copy all other derivatives to a new folder called total_models/derivatives_no_simplification and then run this code.

# Model 4: before simplification
species_list <- df_abundance_re |>
# #   # no simplification
filter(Species %in% c("Alburnus_arborella", "Barbatula_sp_Lineage_I",
                     "Cyprinus_carpio", "Phoxinus_csikii", "Salmo_trutta")) |>
distinct(Species) |> 
pull(Species)

species_list <- sort(species_list)

derivatives <- list()
gam_output <- list()
model_prediction <- list()

for (i in species_list) {
 data <- df_abundance_re |> 
     filter(Species == i)

 gam_output[[i]] <- gam(data = data, Abundance ~ s(mean_last_7days, k = 3) + s(fLake, bs = 're') +  s(fProtocol, bs = 're'), family = ziP())

 lake_data <- species_lake |>
   filter(Species == i)

 lake_list <- distinct(lake_data, Lake) |>
   pull()

 unique_lakes <- distinct(data, Lake) |> 
     pull()

  for (j in lake_list){
    data_lake <- species_lake |>
      filter(Species == i) |>
     filter(Lake == j)

     random_lake <- sample(unique_lakes, 1)

     newdata <- tibble(mean_last_7days = seq(
       from = min(data_lake$temp, na.rm = TRUE),
       to = max(data_lake$temp, na.rm = TRUE), length = 200),
       fLake = factor(random_lake), fProtocol = factor("VERT"))

     derivatives <- derivatives(gam_output[[i]], data = newdata) |>
       mutate(fLake = factor(j)) |>
       mutate(species = factor(i)) |>
       rename(temp = mean_last_7days)

     saveRDS(derivatives, paste0("total_models/derivatives_no_simplification/derivatives_", i, "_",  j, ".rds"))

   }
 }

# save total derivatives as RDS (no exclusions)
#
df_deriv_no_excl <- list.files(path = "total_models/derivatives_no_simplification", pattern = ".rds", full.names = TRUE) |>
  map_dfr(readRDS)
#
saveRDS(df_deriv_no_excl, "total_models/df_deriv_all_no_excl.rds")

```

##Fig. S3ab

```{r}
# cyprinus carpio and lepomis should not be simplified

# Data of the derivatives before simplifying multiple models
# without changes the 5 species were not run with simplified models. 
all_models_derivatives_no_excl <- readRDS("total_models/df_deriv_all_no_excl.rds")
df_deriv_no_excl <- as_tibble(all_models_derivatives_no_excl)

# Plot Figure S3a
#calculating maximum and mean derivatives for each species per lake

df_quantiles <- df_deriv_no_excl |> 
  group_by(species, fLake) |> 
  mutate(max_derivative = max(.derivative)) |> 
  mutate(mean_derivative = mean(.derivative)) |> 
  ungroup()

# calculate quantiles of derivatives
transform(quantile(df_quantiles$.derivative,
                   c(0,0.01,0.05,0.1,0.25,0.5,0.75,0.85,0.9,0.95,0.99,1)))

# add column with the category of quantile each derivative is in
df_quantiles$groups <- cut(df_quantiles$.derivative,              
                     breaks = c(-1.579943e+06, -2.706252e+00, -1.269492e+00,
                                -3.348207e-01, 1.844665e-03, 2.262231e-01,
                                9.864616e-01, 2.018455e+01, 3.013806e+01, 1.648215e+06,
                                1.303406e+07, 1.406101e+07),
                     labels = c("0-1%", "1-5%", "5-10%", "10-25%", "25-50%", "50-75%", "75-85%", "85-90%", "90-95%",
                                "95-99%", "99-100%"))

#  derivatives per lake
plot_percentiles <- df_quantiles |> 
  # Salmo trutta derivative is too large to be plotted
  filter(species != "Salmo_trutta") |> 
  ggplot(aes(x = mean_derivative, fill = factor(species))) + 
  geom_histogram(aes(y = after_stat(count / sum(count))), binwidth = 1.5) +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_manual(values = c("Cyprinus_carpio" = "#80CDC1",
                                "Alburnus_arborella" = "#8C510A",
                                "Barbatula_sp_Lineage_I"="#DFC27D",
                               "Phoxinus_csikii" = "#01665E",  "Salmo_trutta" = "#F46D43", "Cottus_sp_Po_profundal" = "red"))

plot_percentiles

# Plot S3a add lines for the percentiles
plot_percentiles_1 <- plot_percentiles +
  geom_vline(xintercept = 3.013806e+01, color = "red") +
  annotate("text", x= 26, y = 0.12, label="90th percentile", angle=90) +
  geom_vline(xintercept = 2.018455e+01, color = "red") +
  annotate("text", x= 16, y = 0.12, label="85th percentile", angle=90) +
  theme_classic(base_size = 13) +
  xlab("Mean species derivative per lake") +
  ylab("Percentage") +
 guides(fill = guide_legend(
    title = "Species")) +
  ylim(0, 0.2)

plot_percentiles_1

# LOOKING AT MAX DERIVATIVE
# Plot S3a: histogram of species' mean derivatives per lake
df_quantiles |> 
  # Salmo trutta derivative is too large to be plotted
  filter(species != "Salmo_trutta") |> 
  ggplot(aes(x = max_derivative, fill = factor(species))) + 
  geom_histogram(aes(y = after_stat(count / sum(count))), binwidth = 1.5) +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_manual(values = c("Cyprinus_carpio" = "#80CDC1",
                                "Alburnus_arborella" = "#8C510A",
                                "Barbatula_sp_Lineage_I"="#DFC27D",
                               "Phoxinus_csikii" = "#01665E",  "Salmo_trutta" = "#F46D43", "Cottus_sp_Po_profundal" = "red"))

# save as TIFF
# tiff(paste("total_models/plots/plot_deriv_percentiles_1.tiff", sep = ""), units="in", width=12, height=6, res=300)
# 
# plot(plot_percentiles_1)
# 
# dev.off()

# above the 85th percentile

df_quantiles |> 
  filter(.derivative > 2.018455e+01) |> 
  distinct(species)

# 1 Alburnus_arborella    
# 2 Barbatula_sp_Lineage_I
# 3 Cyprinus_carpio       
# 4 Phoxinus_csikii       
# 5 Salmo_trutta

# NEW: upon inclusion of Lepomis gibbosus
# above 90% = 3.013806e+01
df_quantiles |> 
  filter(.derivative > 3.013806e+01) |> 
  distinct(species)

# 1 Alburnus_arborella    
# 2 Barbatula_sp_Lineage_I
# 3 Phoxinus_csikii       
# 4 Salmo_trutta

# 5 species above 90% -> those will be simplified to binomial 
# 1 Alburnus_arborella    
# 2 Barbatula_sp_Lineage_I
# 3 Cyprinus_carpio       
# 4 Phoxinus_csikii       
# 5 Salmo_trutta

# Plotting Figure S3b

# just keeping mean and maximum derivative values for each lake and species
data_quantiles <- df_quantiles |> 
  distinct(mean_derivative, max_derivative, fLake, species)

# Breaking mean derivatives in the same percentiles as before

data_quantiles$percentiles <- cut(data_quantiles$mean_derivative,              
                     breaks = c(-1.579943e+06, -2.706252e+00, -1.269492e+00,
                                -3.348207e-01, 1.844665e-03, 2.262231e-01,
                                9.864616e-01, 2.018455e+01, 3.013806e+01, 1.648215e+06,
                                1.303406e+07, 1.406101e+07),
                     labels = c("0-1%", "1-5%", "5-10%", "10-25%", "25-50%", "50-75%", "75-85%", "85-90%", "90-95%",
                                "95-99%", "99-100%"))

# Plot S3b tile plot of mean derivatives per species in each lake
mean_deriv_plot <- data_quantiles |> 
  ggplot(aes(fLake, y = fct_reorder(species, mean_derivative), fill= percentiles)) + 
  geom_tile() +
  scale_fill_manual(breaks = levels(data_quantiles$percentiles),
                    values = rev(brewer.pal(11, "BrBG")))

derivative_percentiles_2 <- mean_deriv_plot +
  guides(fill = guide_legend(title = "percentiles (mean derivative)", reverse = TRUE)) +
  xlab("") +
  ylab("") +
  theme_classic(base_size = 16)

derivative_percentiles_2

# max derivatives per species in each lake
max_deriv_plot <- data_quantiles |> 
  ggplot(aes(fLake, y = fct_reorder(species, max_derivative), fill= percentiles)) + 
  geom_tile() +
  scale_fill_manual(breaks = levels(data_quantiles$percentiles),
                    values = rev(brewer.pal(11, "BrBG")))

max_deriv_plot +
  guides(fill = guide_legend(title = "percentiles (max derivative)", reverse = TRUE)) +
  xlab("") +
  ylab("") +
  theme_classic(base_size = 16)

# save plot as TIFF
# tiff(paste("total_models/plots/plot_deriv_percentiles_2.tiff", sep = ""), units="in", width=19, height=12, res=300)
# 
# plot(derivative_percentiles_2)
# 
# dev.off()

```

##Fig. S2: Overview derivatives
Overview of derivatives in each lake community

```{r}
# Read in derivative data
all_models_derivatives <- readRDS("total_models/df_deriv_all.rds")
all_derivatives <- as_tibble(all_models_derivatives)

str(all_derivatives)

# Highlighting species with interesting derivatives: 
species_group <- all_derivatives |> 
  filter(species %in% c("Phoxinus_csikii", "Barbatula_sp_Lineage_II", "Cottus_sp_Po_profundal")) |> 
  distinct(species) |> 
  pull(species)

# Plot S2: derivatives included in each lake community
overview_derivatives <- all_derivatives |>
  arrange(fLake) |> 
  ggplot(aes(temp, .derivative, color = species)) +
  geom_line() +
  facet_wrap(~fLake) +
  theme_bw(base_size = 16) +
  scale_color_viridis(discrete = TRUE, breaks = species_group, option = "G") +
  xlab("Temperature") +
  ylab("Derivatives") +
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(), 
        legend.position = "right",
        legend.title = element_blank()) +
  geom_hline(yintercept = 0, color = "#7A7A7A")

overview_derivatives

# save as TIFF
# tiff(paste("total_models/plots/overview_derivatives_lake.tiff", sep = ""), units="in", width=11, height=7, res=300)
# 
# plot(overview_derivatives)
# 
# dev.off()
```

#Calculating thermal response diversity

```{r}
# Function provided by Ross et al. "How to measure response diversity" used to calculate
# response diversity

# read in function
source(here("functions.R"))

# load derivatives data
all_models_derivatives <- readRDS("total_models/df_deriv_all.rds")

all_deriv <- as_tibble(all_models_derivatives)

str(all_deriv)

levels(all_deriv$species)

#Calculating response diversity for each lake community

all_deriv$fLake <- as.character(all_deriv $fLake)

lakes_list <- all_deriv |> 
  distinct(fLake) |> 
  pull(fLake)

str(lakes_list)

lakes_list <- sort(lakes_list)

species_overview <- tibble()

all_deriv$species <- as.character(all_deriv$species)
all_deriv$species <- as.factor(all_deriv$species)

str(all_deriv )

# Loop to get response diversity measures for each lake 
# loop through list of lakes and calculate response diversity in the lake communities

for (i in lakes_list){

  data <- all_deriv |>
    select(.derivative, temp, species, fLake) |> 
    filter(fLake == i)

  df_resp_div <- data |>
    pivot_wider(
      names_from = species,
      values_from = .derivative)

  # apply the function from Ross et al. 
  # Dissimilarity
  df_resp_div$rdiv <- apply(df_resp_div[,-(1:2), drop = FALSE], 1, resp_div, sign_sens = F)
  # Divergence
  df_resp_div$sign <- apply(df_resp_div[,-(1:2), drop = FALSE], 1, resp_div, sign_sens = T)
  # Median Dissimilarity
  df_resp_div$Med <- median(df_resp_div$rdiv)

  # save response diversity metrics per lake
  saveRDS(df_resp_div, paste0("total_models/resp_div_lakes/df_resp_div_", i, ".rds"))

}

# save the response diversity per lake into one file
resp_div_no_excl <- list.files(path = "total_models/resp_div_lakes", pattern = ".rds", full.names = TRUE) |>
  map_dfr(readRDS) |>
  relocate(rdiv, Med, sign, .after = temp)

saveRDS(resp_div_no_excl,"total_models/resp_div_all.rds")
```

>>>>

---
title: "02_models_1"
author: "Sophie Moreau"
date: "2024-09-27"
output: html_document
editor_options: 
  chunk_output_type: console
---
#General Additive Models
Running the predictions for all GAMs with the smaller data frames to reduce running 
time.

@@
